Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-04 MScDiplom_Lemenkova.tex
added 'Index of People'

\documentclass[11pt, a4paper]{article}
\usepackage{authblk}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[a4paper, left=30mm,right=25mm,top=30mm,bottom=35mm]{geometry}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{makeidx}
\makeindex
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

\begin{titlepage}
	\centering
	\title{Seagrass mapping and monitoring \\ along the coasts of Crete, Greece}
	\author[1]{by Lemenkova Polina}
	\affil[1]{\textit{University of Twente, Faculty of Geo-Information Science and Earth Observation (ITC)}, The Netherlands}
	\date{\today}
	\maketitle	
\end{titlepage}
\pagebreak

@@

\begin{itemize}
	\item[*] University of Southampton (UK)\\
	\item[*] Lund University (Sweden)\\
	\item[*] University of Warsaw (Poland)\\
	\item[*] University of Twente, Faculty ITC (The Netherlands)
\end{itemize}

@@

\section*{}
\addcontentsline{toc}{section}{Supervisory Committee}
\begin{center}
\textbf{Seagrass mapping and monitoring along the coasts of Crete, Greece}\\
\vspace{1em}
 {by Lemenkova Polina}
\end{center}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)\dotfill
	\item Name Examiner 2\dotfill
	\item External Examiner \dotfill \\ \hfill{Dr.{\textit{Petter Pilesj\"o}}(Lund University, GIS Centre)}
\end{itemize}

@@

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus} \dotfill \\ (Department of Natural Resource, ITC)
	\item[* ] Dr. \textit{Albertus G. Toxopeus} \dotfill \\ (Department of Natural Resource, ITC)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}  and Mr. \textit{Valentijn Venus} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \author{\textit{Dingtian Yang}} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}\index{Software used}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses\index{Seagrasses} create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species\index{Seagrasses!global species} (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas\index{Seagrasses!distribution of}), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species \index{Seagrasses!Mediterranean species} in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}.

@@

e.g. thermal, sewage, dredging and chemical pollution\index{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,

@@

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index{Seagrasses!natural impacts on} can cause

@@

\subsubsection{General characteristics of \textit{P.oceanica}}\index{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area\index{Seagrasses!\textit{P.oceanica}!distribution} (Fig.\ref{fig:5}), 
with its extension limited by the western part of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphologically\index{Seagrasses!\textit{P.oceanica}!morphology}, \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots and short rhizomes (0.5-2.0 mm). 
The leaves are, perhaps, the most particular characteristics of \textit{P.oceanica}, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:8}(a)): having usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}(c)).
Growing \textit{P.oceanica} make a meadows which, in turn, consist of smaller patches, called \textit{matte}
(Fig.\ref{fig:8}(b)), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.

@@

Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation of the meadow health status. 
There are several environmental factors, determining the growth of the \textit{P.oceanica}. 
The adaptation\index{Seagrasses!\textit{P.oceanica}!environmental adaptation}
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.
The phenology\index{Seagrasses!\textit{P.oceanica}!phenology} of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}\index{Research}
Monitoring\index{Research!problem} of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

Understanding radiance distribution within a water colum\index{Seawater optical properties!RTM}n is necessary for the studies of the underwater visibility, because scattering properties of the water body naturally vary with changing depth, wavelength and environmental conditions.  Simulation of the radiance quantities for natural water bodies enables to analyse seafloor color remote sensing properties. Therefore, the artificial modelling of the seawater optical properties by means of the \ac{RTM} is used when water optics is studied under changing environment (e.g., depths, sun angle, suspended particles in water column). \\ In such cases a retrieval of water optical parameters from the remote measurements should be tested and analyzed. 

@@

		\item [*] The \ac{6SV1} \index{Seawater optical properties!RTM!6SV1} is a US atmospheric correction algorithm, developed in the University of Maryland; adjusted for the \ac{MODIS} satellite imagery lookup tables. The 6S algorithm has been \href{http://grass.osgeo.org/grass64/manuals/html64_user/i.atcorr.html}{implemented} by the \ac{GRASS} software for the atmospheric correction 				 
		\item [*] Another example of the RTM adjusted for a specific imagery is a German \ac{KOPRA}\index{Seawater optical properties!RTM!KOPRA}, fitted for \ac{MIPAS} and \ac{ENVISAT} imagery.

@@

\subsection{Data pre-processing}\index{Data!pre-processing}

@@

Each time we observe the outcome of a random variable\index{Data!statistical analysis!variables} of spectral reflectance, we obtain an observation, which is hence no longer random. 

>>>>

\documentclass[11pt, a4paper]{article}
\usepackage{authblk}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[a4paper, left=30mm,right=25mm,top=30mm,bottom=35mm]{geometry}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

%\newcommand{\listpeoplename}{}
%\newlistof{people}{peo}{\listpeoplename}
%\newcommand{\people}[1]{\refstepcounter{people}}

%\newindex[List of People]{person}
%\setindexpreamble[person]{...}
%\newcommand{\person}[2]{\sindex[person]{#2, #1}#2}% "Given Surname" in text

\phantomsection	
\addcontentsline{toc}{section}{Front Page}
\begin{titlepage}
	\centering
	\title{Seagrass mapping and monitoring \\ along the coasts of Crete, Greece}
	\author[1]{by Lemenkova Polina}
	\affil[1]{MSc student at the \textit{University of Twente, Faculty of Geo-Information Science and Earth Observation (ITC)}, The Netherlands}
	\date{\today}
	\maketitle	
\end{titlepage}
\pagebreak

@@

\begin{itemize}
	\item[*] \href{http://www.southampton.ac.uk/}{University of Southampton}, \href{http://www.southampton.ac.uk/geography/}{School of Geography} (UK)\\
	\item[*] \href{http://www.lunduniversity.lu.se/}{Lund University}, \href{http://www.giscentrum.lu.se/english/index.htm}{GIS Centre} (Sweden)\\
	\item[*] \href{http://www.uw.edu.pl/en/}{University of Warsaw} (Poland)\\
	\item[*] \href{http://www.utwente.nl/en}{University of Twente}, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)} (The Netherlands)
\end{itemize}

@@

\section*{}
\addcontentsline{toc}{section}{Supervisory Committee}
\begin{center}
\textbf{Seagrass mapping and monitoring along the coasts of Crete, Greece}\\
\vspace{1em}
by Lemenkova Polina\index[people]{Lemenkova, P.}
\end{center}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)\dotfill
	\item Name Examiner 2\dotfill
	\item External Examiner \dotfill \\ \hfill{Dr.{\textit{Petter Pilesj\"o}\index[people]{Pilesj\"o, P.}} (Lund University, GIS Centre)}
\end{itemize}

@@

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} \dotfill \\ (Department of Natural Resource, ITC)
	\item[* ] Dr. \textit{Albertus G. Toxopeus}\index[people]{Toxopeus, A.G.} \dotfill \\ (Department of Natural Resource, ITC)
\end{itemize}

@@

This research was carried out at the University of Twente, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)}, Enschede, the Netherlands, under the financial support of the EU \href{http://ec.europa.eu/education/external-relation-programmes/doc72_en.htm}{Erasmus Mundus Scholarship}. 
\vspace{1mm}\\
I thank my scientific supervisors, Dr. \textit{Albertus G.Toxopeus}\index[people]{Toxopeus, A.G.}  and Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} from the ITC, Department of Natural Resources, for our successful scientific collaboration, help and discussions of my research work. \\
I also express them my most sincere appreciations for our fieldwork on Crete. \\ Besides scientific routine, we had an unforgettable time spent together. \\ Personally, I would like to mention that I have visited Greece the first time in my life, and I will always remember my Cretan period-2010 and Ligaria beach.\vspace{1mm}\\
I extend my gratitude to the staff of the Museum of Natural History, the University of Crete, especially to Dr.\textit{Petros Limberakis}\index[people]{Limberakis, Petros} and his colleagues, for their assistance during our fieldwork, providing us with all necessary equipment (e.g. SCUBA gear diving, instruments for underwater measurements, etc). I also appreciate their very warm, personal attitude and interest towards our research. 
\vspace{1mm}\\
My sincere gratitude goes again to Mr.\textit{Valentijn Venus}\index[people]{Venus, V.} for his generous providing me with a lot of spatial data   relevant to my research (e.g. such as digital vector layers of Crete island, raster georeferenced maps for the whole island for my GIS project, Excel spreadsheets, etc). Without these vital data the implementation of this project would have hardly been possible. I am also thankful to him for the technical help in Google Earth imagery capturing, using GDAL technologies and assistance during the fieldwork (videometric underwater footage, consultancies in seagrass observation and monitoring, driving boat, etc). \\ My special thank to his personal, friendly support in various situations on Crete.\vspace{1mm}\\
I am thankful to Dr.rer.nat.\textit{Peter Gege}\index[people]{Gege P.}, the software developer from the Earth Observation Centre of the German Space Agency, for his generous consultancies concerning the technical question about WASI that arose during my work. I am thankful for all his explanations and articles given to me via e-mail \vspace{1mm}\\
My special thanks are addressed to my elder brother, Mr.\textit{Peter Lemenkov}\index[people]{Lemenkov, P.}, the software developer from the ipPort, for his priceless numerous advices and constant professional support throughout my career. He encouraged me to use various very useful software and tools, always gave me necessary recommendations and precious help, thus highly contributing to my professional growth. \\ My special thank for his help with Python scripting. \\
I am also heartfelt thankful to him for his personal friendship and support. 
\vspace{1mm}\\
I am grateful to Mr.\textit{Willem Nieuwenhuis}\index[people]{Nieuwenhuis, W.}, the software developer from the ITC, Department of Natural Resources, for shared technology of Google Earth aerial imagery grabbing and for his consultancy with ASTER imagery. 
\vspace{1mm}\\
I thank Dr. Ir. \textit{Cornelis A.J.M. de Bie}\index[people]{de Bie C.A.J.M.} from the Department of Natural Resource, ITC, for his patient sharing of iPAQ knowledge during my Polish fieldwork in summer 2010. I used this skills on Crete. 
\vspace{1mm}\\
Many thanks for Dr. \textit{Dingtian Yang}\index[people]{Yang D.} from the South China Sea Institute of Oceanology, our colleague in problem of seagrass monitoring and research, for his handing over the data of spectral reflectance of Thalassia seagrass, which enabled to compare spectra of various seagrass types.

@@

\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}\index[concepts]{Software used}

@@

\subsection{Summary}
The seagrasses, a unique group of aquatic plants growing submerged in the sea water, with root-like
structures (rhizoms) buried in the sediments and vertical elongate leaves. A flowering plant,
completely adapted to the marine environment, they are however, more closely related to the lily family
(\textit{Liliaceae}) than to the true grass, despite their name \textit{seagrass}, caused by the ribbon-like, grassy leaves
\cite{McKenzie08}\label{McKenzie08}. Seagrasses\index[concepts]{Seagrasses} create unique, complex, extremely diversified and productive ecological systems in the
littoral coastal zones between 0-50 meters in shallow waters all over the world \cite{Hogarth07}\label{Hogarth07}, and
serve as valuable environmental indicators for the marine ecosystems health. Seagrasses are closely
connected and linked with complex interactions to other vegetation types, e.g. mangroves, coral reefs,
etc. An important constructing component of littoral ecosystems, seagrass contributes significantly to
their structure and functioning.\\
The adaptation to the salt waters is evidently influenced the global distribution of the seagrasses,
limiting it to shallow coastal areas. A number of critical conditions determine growth of the seagrass,
including general climatic characteristics of the area, i.e.temperature, day length, geological and
geomorphological conditions, e.g. soft type of sediments (sand or mud), shallow depths, as well as
chemical and physical parameters of the water: salinity, waves \cite{McKenzie02}\label{McKenzie02}.
The seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}) is a key species to inhabit littoral of the
Mediterranean Sea, see Fig. \ref{fig:1}

@@

\subsection{Background}
\subsubsection{Global distribution of the seagrasses}
Globally, there are 58 recognized and described seagrass species\index[concepts]{Seagrasses!global species} (Fig. \ref{fig:2}), belonging to two orders
(\textit{Hydrocharitales} and \textit{Najadales}), four families (\textit{Hydrocharitaceae, Posidoniaceae, Cymodoceaceae
and Zosteraceae}), and 12 genera (\textit{Enhalus, Thalassia, Halophila, Posidonia, Syringodium, Halodule,
Cymodocea, Amphibolis, Thalassodendron, Zostera, Heterozostera} and \textit{Phyllospandix}) \cite{Kuo89}\label{Kuo89}.

@@

The distribution of the seagrasses is strongly influenced by several environmental factors, which
include climate (mostly, tropical and temperate areas\index[concepts]{Seagrasses!distribution of}), bathymetry (shallow shelf zones), hydrological
particularities (chemical content of water, nutrient availability and turbidity of waves), and geological 
characteristics - sedimentation and cover types of the seafloor \cite{McKenzie06}\label{McKenzie06}.\\
There are four European seagrass species \index[concepts]{Seagrasses!Mediterranean species} in Mediterranean area \cite{Borum04}\label{Borum04}: \textit{Zostera marina,
Zostera noltii, Cymodocea nodosa} and \textit{P.oceanica}. In Greece the common species are \textit{P.oceanica}
\textit{(L.) Delile, Cymodocea nodosa (Ucria) Ascherson, Zostera noltii Hornemann} and \textit{Halophila
stipulacea} \cite{Amoutzopoulou-Schina05}\label{Amoutzopoulou-Schina05}. 
These species differ in morphological and phenological features (Fig. \ref{fig:3}) as well as in structure and dynamics. Thus, \textit{Cymodocea nodosa} is
considered the pioneer species of \textit{P.oceanica} beds, the latter species forming the last stage. When \textit{P.oceanica} beds regresses, \textit{C.nodosa} often replaces them \cite{DenHartog77}\label{DenHartog77}; as a result, \textit{P.oceanica},
\textit{C.nodosa}, and \textit{Z.noltii} do not form mixed persistent stands \cite{Buia91}\label{Buia91}.

@@

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index[concepts]{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}.

@@

e.g. thermal, sewage, dredging and chemical pollution\index[concepts]{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,

@@

Besides anthropogenic factors, various biochemical, climatic and environmental processes\index[concepts]{Seagrasses!natural impacts on} can cause

@@

\subsubsection{General characteristics of \textit{P.oceanica}}\index[concepts]{Seagrasses!\textit{P.oceanica}}
The endemic Mediterranean seagrass \textit{Posidonia oceanica} (further \textit{P.oceanica}), is a main species in
marine coastal environment of Greece, forming, despite its slow growth, the largest, most widespread,
homogeneous and dense meadows in the Mediterranean between 5 and 40 m depth \cite{DenHartog70}\label{DenHartog70}.

@@

The dominant and most productive coastal ecosystem of the Mediterranean, \textit{P.oceanica} is
spatially restricted to the Mediterranean area\index[concepts]{Seagrasses!\textit{P.oceanica}!distribution} (Fig.\ref{fig:5}), 
with its extension limited by the western part of the Mediterranean Sea where cold Atlantic waters enter Gibraltar and mix with warm
Mediterranean waters, thus decreasing its temperature.
Morphologically\index[concepts]{Seagrasses!\textit{P.oceanica}!morphology}, \textit{P.oceanica} consists of long, 5-12 mm broad, “hairy-like” leaves, 3-4 mm thick roots and short rhizomes (0.5-2.0 mm). 
The leaves are, perhaps, the most particular characteristics of \textit{P.oceanica}, making it highly recognizable and distinguishable from other seagrasses (Fig.\ref{fig:8}(a)): having usual length of 20-40 cm, in some cases they can reach up to 1 m \cite{Borum04}\label{Borum04}  (Fig.\ref{fig:8}(c)).
Growing \textit{P.oceanica} make a meadows which, in turn, consist of smaller patches, called \textit{matte}
(Fig.\ref{fig:8}(b)), a monumental construction made by the growth of rhizomes and leaves with entangled roots
and entrapped sediment \cite{Francour06}\label{Francour06}.

@@

Representing one of the most productive Mediterranean ecosystems, \textit{P.oceanica} usually serves as a
perfect biological indicator for the assessment of the quality of waters and environmental health
\cite{Boudouresque89}\label{Boudouresque89}. Some authors \cite{Guidetti08,Montefalcone09}\label{Guidetti08} \label{Montefalcone09} used status and population dynamics of \textit{P.oceanica} as indicators for the evaluation of the meadow health status. 
There are several environmental factors, determining the growth of the \textit{P.oceanica}. 
The adaptation\index[concepts]{Seagrasses!\textit{P.oceanica}!environmental adaptation}
to dry-summer subtropical climate reduces its extension to Mediterranean area only (Fig.\ref{fig:5}).
Besides, the distribution of the seagrasses changes with water depth: it is noticed \cite{Dural10}\label{Dural10} that
the highest flowering density is usually in the 4-7 m depth. 
\textit{P.oceanica} flowers appeared in shallow
stands in September while in November only in stands deeper than 15 m. This time delay is caused by
the different maximum summer temperatures at those depths \cite{Buia91}\label{Buia91}.
The phenology\index[concepts]{Seagrasses!\textit{P.oceanica}!phenology} of \textit{P.oceanica} is also affected by the coastal bathymetry: in the isolated meadows in shallow waters plants have shorter and falciform leaves, compared to ones in the deeper and central
areas \cite{Dural10}\label{Dural10}. In \textit{P.oceanica} flower abundance is related to the structure of the meadow with
the maximal flower density in the densest stands, while the occurrence of flowering is regulated by
environmental factors \cite{Buia91}\label{Buia91}. Phenology of the \textit{P.oceanica} undergoes 
modifications with varying seasons during the year: during the flowering period (ca 3 months long)
the number of leaves on the flowering shoots decrease. Changes of the leaf growth also appear in the
flowering shoots with longer oldest leaves and shorter and narrower leaves induced during the
flowering \cite{Gobert01}\label{Gobert01}.

\subsection{Research problem}\index[concepts]{Research}
Monitoring\index[concepts]{Research!problem} of the marine benthic ecosystems of seagrasses is essential for the environmental
assessment of the coastal zones. It increases our knowledge of the seagrass ecology, highlights 
threats to the seagrass and preventing them from possible losses and degrading and improves
techniques and methods of the underwater-based observations. Mapping the seagrass contributes to
the evaluating of the seagrass current distribution, analysis of its dynamics and changes over time, as
well as estimations of the degrading of seagrass meadows for the purpose of the coastal management.\\
Precise, correct and up-to-date information about the distribution of \textit{P.oceanica} is necessary for the
sustainable conservation of the marine environment and ecosystems in Mediterranean area, being an
important contribution to the environmental coastal zone management \cite{Pergent-Martini06}\label{Pergent-Martini06}.
However, mapping the seagrass has limitations due to the specific location and characteristics of the
research object. The remote sensing techniques have traditionally been widely used for the seagrass
monitoring.

@@

Understanding radiance distribution within a water colum\index[concepts]{Seawater optical properties!RTM}n is necessary for the studies of the underwater visibility, because scattering properties of the water body naturally vary with changing depth, wavelength and environmental conditions.  Simulation of the radiance quantities for natural water bodies enables to analyse seafloor color remote sensing properties. Therefore, the artificial modelling of the seawater optical properties by means of the \ac{RTM} is used when water optics is studied under changing environment (e.g., depths, sun angle, suspended particles in water column). \\ In such cases a retrieval of water optical parameters from the remote measurements should be tested and analyzed. 

@@

		\item [*] The \ac{6SV1} \index[concepts]{Seawater optical properties!RTM!6SV1} is a US atmospheric correction algorithm, developed in the University of Maryland; adjusted for the \ac{MODIS} satellite imagery lookup tables. The 6S algorithm has been \href{http://grass.osgeo.org/grass64/manuals/html64_user/i.atcorr.html}{implemented} by the \ac{GRASS} software for the atmospheric correction 				 
		\item [*] Another example of the RTM adjusted for a specific imagery is a German \ac{KOPRA}\index[concepts]{Seawater optical properties!RTM!KOPRA}, fitted for \ac{MIPAS} and \ac{ENVISAT} imagery.

@@

\subsection{Data pre-processing}\index[concepts]{Data!pre-processing}

@@

Each time we observe the outcome of a random variable\index[concepts]{Data!statistical analysis!variables} of spectral reflectance, we obtain an observation, which is hence no longer random. 
